Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus (Extended Version)

### 2025-04-21 paper/paper.tex
appendix -> extended version

We present a selection of the non-linear type constructors in
\Cref{fig:non-linear-types} and provide the rest in
\ifarxiv{\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of
    \cref{sec:syntax}}\else{the appendix}\fi.  First,
we include universe types $U$ of small non-linear types and $L$ of linear
types. These are defined as
universes ``ala Coquand'' in that we define judgments saying when
non-linear and linear types are \emph{small} and define the universes to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. The definition of
smallness is simply that all types are small as long as their sub-formulae are,
with the exception of the two universe types themselves. A formal
description of smallness is given in
\ifarxiv{\cref{fig:small-nonlin-ty,fig:small-lin-ty} of
    \cref{sec:syntax}}\else{the appendix}\fi. These universe
types are needed so that we can define non-linear and linear types by recursion on natural
numbers.
Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number
types. More complex inductive types --- such as sum types, list types, and
$Fin~n$ --- can be defined in terms of these primitives, and we give their
encoding in \ifarxiv{\cref{sec:syntax}}\else{the appendix}\fi.

@@

We give the rules for linear type
formation in \cref{fig:linear-formation} and the definition of linear terms in
\cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and is included in
\ifarxiv{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the appendix}\fi.

@@

Next, we introduce the most complex and important linear type
constructors of our development, \emph{indexed inductive linear
types}. We encode these by adding a mechanism for constructing initial
algebras of strictly positive functorial type expressions, following
prior work on inductive types
\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
\emph{strictly positive functorial} linear type expressions indexed by
a non-linear type $X$. We think of the elements of this type as
syntactic descriptions of linear types that are parameterized by
$X$-many variables standing for linear types that are only used in
strictly positive positions. Accordingly, the $\SPF\,X$ type supports
an operation $\el$ that interprets it as such a type constructor, as
well as an operator $\map$ that defines a functorial action on parse
transformers. The $\SPF\,X$ type supports constructors for a reference
$\Var~x$ to one of the linear type variables, a constant expression
that doesn't mention any type variables $K$, as well as tensor
products and additive conjunction and disjunction of type expressions.
Further, we add equations in \ifarxiv{\cref{fig:spf-act} of \cref{sec:syntax}
  }\else{the appendix }\fi that say that the
$\el$/$\map$ operations correspond to these descriptions of the
constructors.

@@

The remaining details of the interpretation of linear terms as parse
transformers is a relatively straightforward extension of existing semantics for
linear logic in monoidal categories \cite{seely89}. In
\ifarxiv{\cref{sec:denotational}}\else{the appendix}\fi, we include the denotations
of linear terms and prove that our semantics respects the equational theory of
\theoryabbv --- as well as \cref{ax:dist,ax:disjointness,ax:string-top}.

>>>>

We present a selection of the non-linear type constructors in
\Cref{fig:non-linear-types} and provide the rest in
\ifarxiv{\cref{fig:full-non-linear-types,fig:jdg-eq-nonlinear} of
    \cref{sec:syntax}}\else{the extended version of the paper}\fi.  First,
we include universe types $U$ of small non-linear types and $L$ of linear
types. These are defined as
universes ``ala Coquand'' in that we define judgments saying when
non-linear and linear types are \emph{small} and define the universes to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. The definition of
smallness is simply that all types are small as long as their sub-formulae are,
with the exception of the two universe types themselves. A formal
description of smallness is given in
\ifarxiv{\cref{fig:small-nonlin-ty,fig:small-lin-ty} of
    \cref{sec:syntax}}\else{the extended version of the paper}\fi. These universe
types are needed so that we can define non-linear and linear types by recursion on natural
numbers.
Next, we include standard $\Sigma,\Pi$, empty, unit, Boolean, and natural number
types. More complex inductive types --- such as sum types, list types, and
$Fin~n$ --- can be defined in terms of these primitives, and we give their
encoding in \ifarxiv{\cref{sec:syntax}}\else{the extended version of the paper}\fi.

@@

We give the rules for linear type
formation in \cref{fig:linear-formation} and the definition of linear terms in
\cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and is included in
\ifarxiv{\cref{fig:jdgeq} of \cref{sec:denotational}}\else{the extended version of the paper}\fi.

@@

Next, we introduce the most complex and important linear type
constructors of our development, \emph{indexed inductive linear
types}. We encode these by adding a mechanism for constructing initial
algebras of strictly positive functorial type expressions, following
prior work on inductive types
\cite{nakov_quantitative_2022,altenkirch_indexed_2015}. The syntax is
given in \Cref{fig:iilt}. First, we add a non-linear type $\SPF\,X$ of
\emph{strictly positive functorial} linear type expressions indexed by
a non-linear type $X$. We think of the elements of this type as
syntactic descriptions of linear types that are parameterized by
$X$-many variables standing for linear types that are only used in
strictly positive positions. Accordingly, the $\SPF\,X$ type supports
an operation $\el$ that interprets it as such a type constructor, as
well as an operator $\map$ that defines a functorial action on parse
transformers. The $\SPF\,X$ type supports constructors for a reference
$\Var~x$ to one of the linear type variables, a constant expression
that doesn't mention any type variables $K$, as well as tensor
products and additive conjunction and disjunction of type expressions.
Further, we add equations in \ifarxiv{\cref{fig:spf-act} of \cref{sec:syntax}
  }\else{the extended version of the paper}\fi that say that the
$\el$/$\map$ operations correspond to these descriptions of the
constructors.

@@

The remaining details of the interpretation of linear terms as parse
transformers is a relatively straightforward extension of existing semantics for
linear logic in monoidal categories \cite{seely89}. In
\ifarxiv{\cref{sec:denotational}}\else{the extended version of the paper}\fi, we include the denotations
of linear terms and prove that our semantics respects the equational theory of
\theoryabbv --- as well as \cref{ax:dist,ax:disjointness,ax:string-top}.
